---
title: "Calculating Volumes From Timestamps Using R"

output: 
    html_document:
       css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction

A problem we have been tasked with is forecasting volumes for our Emergency Department (ED). The problem sounds straightforward enough, but it can be formulated in a couple ways

  1. How many patients will be arriving to the ED at some time point in the future (arrival volumes)?
  2. How many patients will be sitting in the ED at some point in the future (patient census)?
  
Depending on the question we wish to tackle we will have to set up our time series data in different formats. Often, the data is presenting to us with raw timestamps of patient arrivals and departures. The following brief tutorial will walk you through setting up data in formats appropriate to answer the questions above using R. 

## The data
  
You can access the simulated data in .csv format from this [github page](https://github.com/LKS-CHART/calculating_ed_volumes), along with all of the code from this example. The data set contains 500 observations with two variables: `arrival_times` representing arrival times and `depart_times`, representing departure times. 


```{r, warning = F, message = F}

# load the required libraries
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

# import the data
ed_data <- read_csv('simulated_ed_data.csv') %>% 
  print()


```


## Calculating Arrival Volumes

Arrival volumes can be calculated by counting the number of rows of data within a specified time window (hour, day, week, etc...). Functions from the `lubridate` package make this an especially straight forward task. Below, we will calculate the number of arrivals  per hour  and plot the results using `ggplot`

```{r, echo}
# calculate volumes per minute
volumes_hour <- ed_data %>% 
  group_by(year = year(arrival_times),
           month = month(arrival_times),
           day = day(arrival_times),
           hour = hour(arrival_times)) %>% 
  summarise(volume = n()) %>% 
  mutate(timestamp = ymd_h(paste(year, month, day, hour))) %>% 
  print()

volumes_hour %>% 
  ggplot(aes(timestamp, volume)) + 
  geom_line() + 
  xlab('date') +
  ggtitle('Emergent Department Arrivals Per Hour') +
  coord_cartesian(ylim = c(0, 20)) +
  scale_y_continuous(breaks = seq(0,20, 2))

```


It may be the case that some intervals in your data won't have any observations (e.g. No arrivals at 2am). In that scenario, you will need to fill in the holes of your time series. The code below shows one solution to this problem for calculating the number of arrivals per minute. You can simply merge your calculated volumes data with another dataset that contains all of the timestamps within the time window of interest. 

```{r}

volumes_minute <- ed_data %>% 
  group_by(year = year(arrival_times),
           month = month(arrival_times),
           day = day(arrival_times),
           hour = hour(arrival_times),
           minute = minute(arrival_times)) %>% 
  summarise(volume = n()) %>% 
  mutate(timestamp = ymd_hm(paste(year, month, day, hour, minute))) %>% 
  ungroup() %>% 
  select(timestamp, volume) %>% 
  print()

# note there are no arrivals between  00:05 and 00:15

# create a sequence of times from the start to end of your available data
start <- min(volumes_minute$timestamp)
end   <- max(volumes_minute$timestamp)
full_time_window <- tibble(timestamp = seq(start, end,
                                           by = 'mins'))

# do a right join to get all of the timestamps

volumes_minute <- volumes_minute %>% 
  right_join(full_time_window, by = 'timestamp') %>% 
  mutate(volume = ifelse(is.na(volume), 0, volume)) %>%  # replace the missing volumes with a 0
  print()

volumes_minute %>% 
  ggplot(aes(timestamp, volume)) + 
  geom_line() + 
  ggtitle('Emergent Department Arrivals Per Minute') +
  coord_cartesian(ylim = c(0, 3)) +
  xlab('date')

```


## Calculating Census Volumes

Calculating census volumes, that is the number of patients waiting in the ED at any given time, requires a different strategy. Among the many strategies of accomplishing this, I will focus on one of the faster algorithms. 

The strategy is to imagine having a counter that keeps track of every time a patient enters or leaves the ED. When a patient walks in the door we add one to the overall count, and when a patient leaves we substract one. Since we have the timestamps of when patients enter and leave, this is a very simple task. 

We will split our data into two chunks, one for arrivals and one for departures. For each data split we will create a counter variable. This variable will take the value of 1 for the arrival split and -1 for the departure split. We then bind the two splits back together, arrange them by time, and take a cummulative sum of the counter variable. 

As with the previous example we need to fill in the gaps where no arrivals or departures exist. Except this time, instead of filling in the gaps with zeros we take the last observation carried forward. Since the data will be arranged by time we can use `tidyr::fill()`. 


```{r}
ed_data <- ed_data %>% 
  mutate(arrival_times = floor_date(arrival_times, unit = 'minute'),
         depart_times = floor_date(depart_times, unit = 'minute'))
arrivals <- ed_data %>% 
  select(timestamp = arrival_times) %>% 
  mutate(counter = 1)

departures <- ed_data %>% 
  select(timestamp = depart_times) %>% 
  mutate(counter = -1)


census_volumes <- arrivals %>% 
  bind_rows(departures) %>% 
  arrange(timestamp, counter) %>%  # arrange by time
  mutate(volume = cumsum(counter)) # take cumulative sum of the counter

# create a sequence of times from the start to end of your available data
start <- min(census_volumes$timestamp)
end   <- max(census_volumes$timestamp)
full_time_window <- tibble(timestamp = seq(start, end,
                                           by = 'mins'))
census_volumes <- census_volumes %>% 
  right_join(full_time_window, 
             by = 'timestamp') %>% # bind with the full time window to fill gaps
  arrange(timestamp) %>% 
  fill(volume, .direction = 'down') # take last observation carried forward

census_volumes
```



```{r}
census_volumes %>% 
  ggplot(aes(timestamp, volume)) + 
  geom_line() + 
  xlab('Date') + 
  ggtitle('Emergency Department Census',
          subtitle = 'In one minute intervals')
```


It will rarely be the case where you will be calculating these volumes from scratch, i.e. from the opening of the ED. Let's redo the above example but take October 2nd at midnight as our starting time.

We need to set a start time for our time window. We then calculate the number of people in the ED at the begining of our time window. We preceed as before, but before taking the cumulative sum we set the first row of the data to be the number of people in the ED at the start time.

```{r}
# set a start time
start_time <- ymd_hms('2018-10-02 00:00:00')


# get the base count
# This is the number of people who arrived to the ED before the start time
# but are still in the ED after the start time
pre <- ed_data %>%
    filter(arrival_times < start_time &
           depart_times >= start_time)

# Anyone who leaves the ED after the start_time
post <- ed_data %>%
  filter(depart_times >= start_time)
        

# The arrivals
arrivals <- post %>% 
  select(timestamp = arrival_times) %>% 
  mutate(counter = 1)%>%
  filter(timestamp >= start_time)

# The departures
departures <- post %>% 
  select(timestamp = depart_times) %>% 
  mutate(counter = -1)

  
census_volumes <- arrivals %>% 
  bind_rows(departures) %>% 
  arrange(timestamp, counter) 

#  Add the starting pre volume to the first row
census_volumes$counter[1] <- census_volumes$counter[1] + nrow(pre)
  
census_volumes <- census_volumes %>%  # arrange by time
  mutate(volume = cumsum(counter)) 

# create a sequence of times from the start to end of your available data
start <- min(census_volumes$timestamp)
end   <- max(census_volumes$timestamp)
full_time_window <- tibble(timestamp = seq(start, end,
                                           by = 'mins'))
census_volumes <- census_volumes %>% 
  right_join(full_time_window, 
             by = 'timestamp') %>% # bind with the full time window to fill gaps
  arrange(timestamp) %>% 
  fill(volume, .direction = 'down') # take last observation carried forward

census_volumes


```


```{r}
census_volumes %>% 
  ggplot(aes(timestamp, volume)) + 
  geom_line() + 
  xlab('Date') + 
  ggtitle('Emergency Department Census',
          subtitle = 'In one minute intervals')
```

## Putting census volume calculation into a function

Since this is a data formating step that we may run repeatedly, it may be a good
idea to create a function to calculate census volumes. The functions handles 
potential timestamp issues (end times < start times, and missing times).
All packages used in the function are explicitly stated using the
with `::` operator. 

```{r}

calc_census_volume <- function(data, start_var, end_var, 
                                 begin_time = '2018-10-02 00:00:00') {
  sv <- rlang::enquo(start_var)
  ev <- rlang::enquo(end_var)
  
   if(!inherits(begin_time, 'POSIXct')) {
    begin_time <- lubridate::ymd_hms(begin_time)
  }

  
  # keep track of timestamp issues
  incorrect_times <- data %>%
    dplyr::filter(!!ev < !!sv | is.na(!!ev) | is.na(!!sv)) %>%
    dplyr::mutate(type = ifelse( is.na(!!ev) ,
                                 'missing end timestamp',
                          ifelse(is.na(!!sv), 
                                 'missing start timestamp', 
                                 'end timestamp before start timestamp')))
  # remove timestamp problems
  filtered_data <- data %>%
    dplyr::filter(!!sv <= !!ev) %>%   # no end timestamps after start timestapms
    dplyr::filter(!is.na(!!ev)) %>%  # no missing end times
    dplyr::filter(!is.na(!!sv))  # no missing start times
    
  
  # get the base count
  pre <- filtered_data %>%
    dplyr::filter(!!sv < begin_time &
             !!ev >= begin_time)
  
   post <- filtered_data %>%
    dplyr::filter(!!ev >= begin_time)
   
  # The arrivals
  arrivals <- post %>% 
    dplyr::select(timestamp = !!sv) %>% 
    dplyr::mutate(counter = 1) %>%
    dplyr::filter(timestamp >= start_time)
  
  # The departures
  departures <- post %>% 
    dplyr::select(timestamp = !!ev) %>% 
    dplyr::mutate(counter = -1)
  
  # bind the arrivals and departures
  census_volumes <- arrivals %>% 
    dplyr::bind_rows(departures) %>% 
    dplyr::arrange(timestamp, counter) 

  # Add the starting pre volume to the first row
  census_volumes$counter[1] <- census_volumes$counter[1] + nrow(pre)
  
  census_volumes <- census_volumes %>%  # arrange by time
    dplyr::mutate(volume = cumsum(counter)) 

  # create a sequence of times from the start to end of your available data
  start <- min(census_volumes$timestamp)
  end   <- max(census_volumes$timestamp)
  full_time_window <- dplyr::tibble(timestamp = seq(start, end,
                                           by = 'mins'))
  census_volumes <- census_volumes %>% 
    right_join(full_time_window, 
               by = 'timestamp') %>% # bind with the full time window to fill gaps
    arrange(timestamp) %>% 
    fill(volume, .direction = 'down') # take last observation carried forward
  
  if(nrow(incorrect_times) > 0) {
    incorrect_times_sum <- incorrect_times %>%
      dplyr::count(type)

    msg <- paste0(paste(incorrect_times_sum$type,
                        incorrect_times_sum$n, sep = ': n = '),
                  collapse = '\n')
    msg <- paste0('Potential Data Issues:\n',
                  msg, 
                  '\nThese timestamp errors have been removed from the 
                  data prior to volume calculation')
    message(msg)
  }

  return(census_volumes)

}
                                 
# try the function

ed_data %>% 
  calc_census_volume(start_var = arrival_times,
                     end_var = depart_times)
```


## Conclusion

The two strategies discussed here can be used for most volume calculations. Before proceeding with either algorithm it is good practice to have a look at your timestamps for accuracy. In using raw timestamp from an Electronic Medical Records system, we may run into some data inconsistencies. It is always a good idea to run some data quality checks before using such data. For example, make sure that there are no departure timestamps that occur before the arrival timestamps. 
